---
title: "ANUBIS Damage Profile Visualization"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: hide
---

# ANUBIS Damage Profile Visualization

Two analysis modes are supported:

-   per_sgb: one SGB across multiple samples
-   per_sample: multiple SGBs within one sample

For each mode, the script generates two plots:

1.  single-line damage profiles for each grouping unit
2.  mean damage profile with ± SD

## Usage examples

```{r example-per-sgb, eval=FALSE} 
analysis_mode <- "per_sgb" 
selected_sgb <- "SGB12664"}
```

```{r example-per-sample, eval=FALSE} 
analysis_mode <- "per_sample" 
selected_sample <- "Farrer2021_101"}
```

## Output

The script creates two PNGs for each target:

-   target_damage_profiles.png
-   target_damage_mean_sd.png

------------------------------------------------------------------------

## 1) Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(patchwork)

theme_set(theme_bw(base_size = 13))
```

## 2) User settings

```{r settings}
# Input file containing the merged ANUBIS table
input_file <- "merged_anubis_table.tsv"

# Choose one mode:
#   "per_sgb"    => one SGB across multiple samples
#   "per_sample" => one sample across multiple SGBs
analysis_mode <- "per_sample"

# Target SGB used when analysis_mode == "per_sgb"
selected_sgb <- "SGB1234"

# Target sample used when analysis_mode == "per_sample"
selected_sample <- "Sample_name"

# Number of positions shown from each read end
max_position <- 29

# Folder for saved plots
output_dir <- "plots"
```

## 3) Load and validate input

```{r load-data, message=FALSE}
anubis <- read_tsv(input_file, show_col_types = FALSE)

required_columns <- c("sample", "sgb_id")
missing_columns <- setdiff(required_columns, colnames(anubis))

if (length(missing_columns) > 0) {
  stop(
    paste0(
      "The following required columns are missing: ",
      paste(missing_columns, collapse = ", ")
    )
  )
}

valid_modes <- c("per_sgb", "per_sample")

if (!analysis_mode %in% valid_modes) {
  stop(
    paste0(
      "Invalid analysis mode: '", analysis_mode, "'. Choose one of: ",
      paste(valid_modes, collapse = ", ")
    )
  )
}
```

## 4) Select target data

```{r select-data}
if (analysis_mode == "per_sgb") {
  if (!selected_sgb %in% anubis$sgb_id) {
    stop(paste0("SGB '", selected_sgb, "' was not found in the input table."))
  }

  selected_data <- anubis %>% filter(sgb_id == selected_sgb)
  target_name <- selected_sgb
  group_column <- "sample"

} else {
  if (!selected_sample %in% anubis$sample) {
    stop(paste0("Sample '", selected_sample, "' was not found in the input table."))
  }

  selected_data <- anubis %>% filter(sample == selected_sample)
  target_name <- selected_sample
  group_column <- "sgb_id"
}

cat(
  "\nAnalysis mode: ", analysis_mode,
  "\nTarget: ", target_name,
  "\nRows selected: ", nrow(selected_data),
  "\n",
  sep = ""
)
```

## 5) Prepare damage data

ANUBIS reports the two read-terminal damage types used here:

-   5' C→T substitutions
-   3' G→A substitutions

```{r prepare-damage}
prepare_damage <- function(data, mutation) {
  prefix <- ifelse(mutation == "CtoT", "CtoT-", "GtoA-")

  data %>%
    select(sample, sgb_id, starts_with(prefix)) %>%
    pivot_longer(
      cols = starts_with(prefix),
      names_to = "position",
      values_to = "damage"
    ) %>%
    mutate(position = as.numeric(sub(prefix, "", position)))
}

ctot <- prepare_damage(selected_data, "CtoT")
gtoa <- prepare_damage(selected_data, "GtoA")
```

## 6) Plot functions

```{r plot-functions}
plot_profile <- function(data, end, ymax, group_column, show_legend = FALSE) {
  if (end == "5") {
    p <- ggplot(
      data,
      aes(x = position, y = damage, group = .data[[group_column]], colour = .data[[group_column]])
    ) +
      geom_line(linewidth = 0.9, alpha = 0.7) +
      scale_x_continuous(limits = c(0, max_position), breaks = seq(0, max_position, by = 5)) +
      scale_y_continuous(limits = c(0, ymax)) +
      theme_bw(base_size = 15) +
      labs(
        title = "5' C→T",
        x = "Distance from 5' end (bp)",
        y = "Misincorporation frequency",
        colour = NULL
      ) +
      theme(
        legend.position = ifelse(show_legend, "right", "none"),
        panel.grid.minor = element_blank(),
        plot.title = element_text(hjust = 0.5)
      )

  } else {
    p <- ggplot(
      data,
      aes(x = position, y = damage, group = .data[[group_column]], colour = .data[[group_column]])
    ) +
      geom_line(linewidth = 0.9, alpha = 0.7) +
      scale_x_reverse(limits = c(max_position, 0), breaks = seq(0, max_position, by = 5)) +
      scale_y_continuous(limits = c(0, ymax)) +
      theme_bw(base_size = 15) +
      labs(
        title = "3' G→A",
        x = "Distance from 3' end (bp)",
        y = NULL,
        colour = NULL
      ) +
      theme(
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = ifelse(show_legend, "right", "none"),
        panel.grid.minor = element_blank(),
        plot.title = element_text(hjust = 0.5)
      )
  }

  p
}

plot_mean_sd <- function(individual_data, average_data, end, ymax) {
  if (end == "5") {
    p <- ggplot() +
      geom_line(
        data = individual_data,
        aes(x = position, y = damage, group = .data[[group_column]]),
        colour = "grey70",
        alpha = 0.5,
        linewidth = 0.6
      ) +
      geom_ribbon(
        data = average_data,
        aes(x = position, ymin = pmax(mean_damage - sd_damage, 0), ymax = mean_damage + sd_damage),
        fill = "steelblue",
        alpha = 0.25
      ) +
      geom_line(
        data = average_data,
        aes(x = position, y = mean_damage),
        colour = "steelblue",
        linewidth = 1.4
      ) +
      geom_point(
        data = average_data,
        aes(x = position, y = mean_damage),
        colour = "steelblue",
        size = 2
      ) +
      scale_x_continuous(limits = c(0, max_position), breaks = seq(0, max_position, by = 5)) +
      scale_y_continuous(limits = c(0, ymax)) +
      theme_bw(base_size = 15) +
      labs(
        title = "5' C→T",
        x = "Distance from 5' end (bp)",
        y = "Misincorporation frequency"
      ) +
      theme(
        legend.position = "none",
        panel.grid.minor = element_blank(),
        plot.title = element_text(hjust = 0.5)
      )

  } else {
    p <- ggplot() +
      geom_line(
        data = individual_data,
        aes(x = position, y = damage, group = .data[[group_column]]),
        colour = "grey70",
        alpha = 0.5,
        linewidth = 0.6
      ) +
      geom_ribbon(
        data = average_data,
        aes(x = position, ymin = pmax(mean_damage - sd_damage, 0), ymax = mean_damage + sd_damage),
        fill = "steelblue",
        alpha = 0.25
      ) +
      geom_line(
        data = average_data,
        aes(x = position, y = mean_damage),
        colour = "steelblue",
        linewidth = 1.4
      ) +
      geom_point(
        data = average_data,
        aes(x = position, y = mean_damage),
        colour = "steelblue",
        size = 2
      ) +
      scale_x_reverse(limits = c(max_position, 0), breaks = seq(0, max_position, by = 5)) +
      scale_y_continuous(limits = c(0, ymax)) +
      theme_bw(base_size = 15) +
      labs(
        title = "3' G→A",
        x = "Distance from 3' end (bp)",
        y = NULL
      ) +
      theme(
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none",
        panel.grid.minor = element_blank(),
        plot.title = element_text(hjust = 0.5)
      )
  }

  p
}
```

## 7) Summary statistics and plotting limits

```{r summary-statistics}
average_ctot <- ctot %>%
  group_by(position) %>%
  summarise(
    mean_damage = mean(damage, na.rm = TRUE),
    sd_damage = sd(damage, na.rm = TRUE),
    .groups = "drop"
  )

average_gtoa <- gtoa %>%
  group_by(position) %>%
  summarise(
    mean_damage = mean(damage, na.rm = TRUE),
    sd_damage = sd(damage, na.rm = TRUE),
    .groups = "drop"
  )

ymax_individual <- max(c(ctot$damage, gtoa$damage), na.rm = TRUE)
ymax_average <- max(
  c(
    ctot$damage,
    gtoa$damage,
    average_ctot$mean_damage + average_ctot$sd_damage,
    average_gtoa$mean_damage + average_gtoa$sd_damage
  ),
  na.rm = TRUE
)
```

## 8) Single-line damage profiles

```{r single-line-plot}
left_single <- plot_profile(ctot, end = "5", ymax = ymax_individual, group_column = group_column, show_legend = FALSE)
right_single <- plot_profile(gtoa, end = "3", ymax = ymax_individual, group_column = group_column, show_legend = TRUE)

single_line_plot <- (left_single | right_single) +
  plot_annotation(title = paste("Damage profiles:", target_name))

single_line_plot
```

## 9) Mean ± SD plots

```{r mean-sd-plot}
left_mean <- plot_mean_sd(ctot, average_ctot, end = "5", ymax = ymax_average)
right_mean <- plot_mean_sd(gtoa, average_gtoa, end = "3", ymax = ymax_average)

mean_sd_plot <- (left_mean | right_mean) +
  plot_annotation(title = paste("Mean ± SD damage profile:", target_name))

mean_sd_plot
```

## 10) Save figures

```{r save-plots}
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

single_file <- file.path(output_dir, paste0(target_name, "_damage_profiles.png"))
mean_file <- file.path(output_dir, paste0(target_name, "_damage_mean_sd.png"))

ggsave(filename = single_file, plot = single_line_plot, width = 12, height = 6, dpi = 300)
ggsave(filename = mean_file, plot = mean_sd_plot, width = 12, height = 6, dpi = 300)

cat(
  "\nSaved plots:\n\n",
  single_file,
  "\n",
  mean_file,
  "\n",
  sep = ""
)
```
